Uncovering underlying high-level musical content in the time domain \\ \small Leveraging self-supervised deep neural networks, inductive bias, and aural skills to learn deep audio embeddings with applications to MIR downstream tasks

### 2023-06-08 abstract/abstract.tex
Updates from Overleaf

This thesis posits the existence of invariant high-level musical concepts that persist through changes in sonic qualities, akin to the symbolic domain where essence endures despite varying interpretations through different performances, instruments, and styles, among many other variables.

In collaboration with Epidemic Sound AB and the Music Technology Group (MTG) at Universitat Pompeu Fabra (UPF), we used self-supervised contrastive learning to uncover the underlying structure of Western tonal music. We applied deep convolutional neural networks and a triplet loss function, identifying abstract and semantic musical elements without relying on their sonic qualities. This way, we replaced traditional acoustic features with deep audio embeddings, paving the way for high-level, sound-agnostic, and content-sensitive music identification.

Our cognitively-based approach for learning embeddings focuses on using full-resolution data and preserving high-level musical information. Drawing upon our domain expertise, we developed robust transformations to encode heuristic musical concepts that should remain constant. This novel approach aims to reconcile music and machine learning, enhancing machine listening models' efficacy. Preliminary results suggest that our musically-informed technique has significant potential for boundary detection tasks and possibly nearly all MIR downstream tasks that are not purely sonic-based. Consequently, music-motivated embeddings appear promising and potentially adaptable to other tasks constrained by data scarcity.

>>>>

This thesis posits the existence of invariant high-level musical concepts that persist regardless of changes in sonic qualities, akin to the symbolic domain where essence endures despite varying interpretations through different performances, instruments, and styles, among many other, almost countless variables.

In collaboration with Epidemic Sound AB and the Music Technology Group (MTG) at Universitat Pompeu Fabra (UPF), we used self-supervised contrastive learning to uncover the underlying structure of Western tonal music. We applied deep convolutional neural networks and a triplet loss function, identifying abstract and semantic musical elements without relying on their sonic qualities. This way, we replaced traditional acoustic features with deep audio embeddings, paving the way for high-level, sound-agnostic, and content-sensitive music representation.

Our cognitively-based approach for learning embeddings focuses on using full-resolution data and preserving high-level musical information. Drawing upon our domain expertise, we developed robust transformations to encode heuristic musical concepts that should remain constant. This novel approach aims to reconcile music and machine learning, enhancing machine listening models' efficacy. Preliminary results suggest that our musically-informed technique has significant potential for boundary detection tasks and possibly nearly all MIR downstream tasks that are not purely sonic-based. While it didn't outperform state-of-the-art results, it showed very competitive results and room for improvement. Consequently, music-motivated audio embeddings appear promising and potentially adaptable to other tasks constrained by data scarcity.

### 2023-06-08 acknowledgements/acknowledgements.tex
Updates from Overleaf

I want to express my sincerest gratitude to...

\begin{itemize}
\item Carlos Lordelo, and Carl Thomé for their invaluable guidance and support throughout my thesis. Your expertise in MIR helped shape the direction of my research and make it a success. Your unwavering support and encouragement helped me overcome any difficulties I encountered during the process. I also appreciate the time and effort you invested in mentoring me. Your dedication to your students is truly inspiring, and I am honored to have had the opportunity to work with you.
\vspace*{3mm}
\item My co-supervisor
\vspace*{3mm}
\item The Music Technology Group (MTG)
\vspace*{3mm}
\item My colleague at My Sheet Music Transcriptions, Oriol López, for his exceptional support and flexibility throughout my thesis. His understanding and willingness to accommodate my academic commitments effectively allowed me to balance my work and research responsibilities. His guidance and advice on professional matters have been instrumental in my growth as a professional. I am genuinely grateful for his unwavering support and encouragement and for being a great mentor.
\vspace*{3mm}
Thank you all for being excellent advisors and playing a crucial role in completing my thesis. I sincerely appreciate everything you have done for me and will always cherish the knowledge and experience I have gained under your guidance.
\vspace*{3mm}
\item My family
\vspace*{3mm}
\item My girlfriend
\vspace*{3mm}
\end{itemize}

>>>>

I want to extend my deepest gratitude to...

\begin{itemize}
\item Carl Thomé and Carlos Lordeloª, whose expertise in MIR, ML, and DSP has been pivotal to the progress and success of my thesis. Their invaluable guidance and unwavering encouragement have empowered me to overcome the challenges encountered throughout this journey. Their dedication and commitment to mentorship are truly inspiring, and their influence has left an indelible mark on my academic path. The opportunity to work under their mentorship has been a privilege and an honor. Their profound impact is beyond what words can encapsulate.

\vspace*{3mm}
\item Cody Hesse and Sebastian Löf

\vspace*{3mm}
\item Epidemic Sound

\vspace*{3mm}
\item The Music Technology Group (MTG) and Universitat Pompeu Fabra (UPF). My profound thanks go to Xavier Serra for his faith in accepting a non-traditional profile to the master's program. Additionally, I appreciate Pedro Ramoneda and Genís Plaja for their patient assistance throughout my journey.

\vspace*{3mm}
\item My colleague at My Sheet Music Transcriptions, Oriol López, for his exceptional support and flexibility throughout my thesis. His understanding and willingness to accommodate my academic commitments effectively allowed me to balance my work and research responsibilities.

\vspace*{3mm}
Thank you all for being excellent advisors and playing a crucial role in completing my thesis. I sincerely appreciate everything you have done for me and will always cherish the knowledge and experience I have gained under your guidance.

\vspace*{3mm}
\end{itemize}

### 2023-06-08 evaluation/evaluation.tex
Updates from Overleaf

\section{\textit{Embeddiogram} feature}

The \textit{embeddiogram} is derived by applying our pre-trained neural network model to sliding windowed segments across the audio signal, generating a sequence of embedding vectors. These relatively high-dimensional embedding vectors collectively form a two-dimensional description of the audio signal's musical content.

Below is a detailed explanation of how we compute the \textit{embeddiogram} from a given audio signal. This process comprises five key steps, including loading the audio data, slicing the audio data into windowed segments, processing each window using a pre-trained model to produce an embedding, collecting these embeddings, and finally normalizing the Embeddiogram.

@@

\begin{figure}[ht]
  \centering
  \begin{minipage}[b]{.8\linewidth}
    \centering
    \includegraphics[width=\linewidth]{figures/images/salami_391_embeddiogram.png}
    \caption[]{Caption for image1}
    \label{fig:image1}
  \end{minipage}

  \begin{minipage}[b]{.8\linewidth}
    \centering
    \includegraphics[width=\linewidth]{figures/images/samali_391_novelty_curve_peaks.png}
    \caption[]{Caption for image2}
    \label{fig:image2}
  \end{minipage}
\end{figure}

\begin{figure}[ht]
    \centering
    \begin{minipage}{0.45\textwidth}
        \centering
        \includegraphics[width=0.9\textwidth]{figures/images/salami_391_R matrix after Gaussian smoothing.png} % first figure itself
        \caption[]{first figure}
    \end{minipage}\hfill
    \begin{minipage}{0.45\textwidth}
        \centering
        \includegraphics[width=0.9\textwidth]{figures/images/salami_391_Lag matrix after Gaussian smoothing.png} % second figure itself
        \caption[]{second figure}
    \end{minipage}
    \vfill
    \begin{minipage}{0.45\textwidth}
        \centering
        \includegraphics[width=0.9\textwidth]{figures/images/salami_391_Q_matrix.png} % third figure itself
        \caption[]{third figure}
    \end{minipage}\hfill
    \begin{minipage}{0.45\textwidth}
        \centering
        \includegraphics[width=0.9\textwidth]{figures/images/salami_391_binary transitive matrix.png} % fourth figure itself
        \caption[]{fourth figure}
    \end{minipage}
\end{figure}

>>>>

\section{The \textit{embeddiogram}, an audio feature embedding representation}

The \textit{embeddiogram} is derived by applying our pre-trained neural network model to sliding windowed segments across the audio signal, generating a sequence of embedding vectors. These relatively low-dimensional embedding vectors collectively form a two-dimensional description of the audio signal's musical content.

Below is a detailed explanation of how we compute the \textit{embeddiogram} from a given audio signal of length $N$. This process comprises loading the audio data, slicing the audio data into windowed segments, processing each window using a pre-trained model to produce an embedding, collecting and stacking these embeddings, and finally normalizing the \textit{embeddiogram}.

@@

\begin{figure}[ht]
  \centering
  \begin{minipage}[b]{1.0\linewidth}
    \centering
    \includegraphics[width=\linewidth]{figures/images/salami_391_embeddiogram.png}
    \caption[Embeddiogram. Track 391 (SALAMI dataset).]{Embeddiogram. Track 391 (SALAMI dataset).}
    \label{fig:image1}
  \end{minipage}

  \begin{minipage}[b]{1.0\linewidth}
    \centering
    \includegraphics[width=\linewidth]{figures/images/samali_391_novelty_curve_peaks.png}
    \caption[Novelty curve + peaks. Track 391 (SALAMI dataset).]{Novelty curve + peaks. Track 391 (SALAMI dataset).}
    \label{fig:image2}
  \end{minipage}
\end{figure}

\begin{figure}[ht]
    \centering
    \begin{minipage}{0.45\textwidth}
        \centering
        \includegraphics[width=0.9\textwidth]{figures/images/salami_391_R matrix after Gaussian smoothing.png} % first figure itself
        \caption[Track 391 (SALAMI dataset) Self-similarity matrix]{Self-similarity matrix. Track 391 (SALAMI dataset).}
    \end{minipage}\hfill
    \begin{minipage}{0.45\textwidth}
        \centering
        \includegraphics[width=0.9\textwidth]{figures/images/salami_391_Lag matrix after Gaussian smoothing.png} % second figure itself
        \caption[Track 391 (SALAMI dataset) Self-similarity lag matrix]{Self-similarity lag matrix. Track 391 (SALAMI dataset).}
    \end{minipage}
    \vfill
    \begin{minipage}{0.45\textwidth}
        \centering
        \includegraphics[width=0.9\textwidth]{figures/images/salami_391_Q_matrix.png} % third figure itself
        \caption[Track 391 (SALAMI dataset). Q-matrix]{Cumulative matrix. Track 391 (SALAMI dataset).}
    \end{minipage}\hfill
    \begin{minipage}{0.45\textwidth}
        \centering
        \includegraphics[width=0.9\textwidth]{figures/images/salami_391_binary transitive matrix.png} % fourth figure itself
        \caption[Track 391 (SALAMI dataset). Transitive Binary Similarity Matrix]{Transitive Binary Matrix. Track 391 (SALAMI dataset).}
    \end{minipage}
\end{figure}

### 2023-06-08 future work/future work.tex
Updates from Overleaf

hybrid autoencoder architecture + UNet stuff
transforms
stems
different takes
k-fold cross validation

@@

>>>>

The question as per \cite{Turian2022HEAR:Representations} or \cite{Li2023MERT:Training} remains if such general-purpose audio representation can mimic human hearing.

transforms
stems
different takes
k-fold cross validation
hyper-params: kernel size 0.005 times the sample rate

@@

Listening evaluation
2D or 3D latent space visualization as per \ref{fig:manifold}

\input{figures/neural networks/manifold}

### 2023-06-08 introduction/Subsections/Motivation.tex
Updates from Overleaf

>>>>

The MIR community appears to slowly converge and bridge the gap in musicology, theory, perception, and technical complexities by addressing. For example, by studing intermediate-level musical patterns, or \textit{mesostructures}. Deep learning applications in music often fail to consider these mesostructures, akin to Schenkerian analysis' middle ground. As noted by \cite{Mesostructures2023}, attending to these patterns could greatly enhance tasks like music analysis, composition, and retrieval by offering a more comprehensive understanding of music, echoing insights from Schenkerian analysis \cite{Introduction_to_Schenkerian_Analysis}.

### 2023-06-08 introduction/Subsections/Objectives.tex
Updates from Overleaf

The study aims to develop a unified numerical understanding (or embedding \footnote{A learned representation or embedding is a numerical output -usually a fixed-size vector- produced by a machine-learning model from a low-level input. Good representations should be versatile across various tasks and require limited supervision \cite{Turian2022HEAR:Representations}.}) for a music piece that integrates high-level elements like melody contour, harmony relationships, and other comprehensive components such as the relationships between elements unfolding through time, arranging the embedding space to emphasize the clustering of these sophisticated musical content without considering sonic qualities or attributes.

>>>>

The study aims to develop a unified numerical understanding (or embedding \footnote{A learned representation or embedding is a numerical output -usually a fixed-size vector- produced by a machine-learning model. Good representations should be versatile across various tasks and require limited supervision \cite{Turian2022HEAR:Representations}.}) for a music piece that integrates high-level elements like melody contour, harmony relationships, and other comprehensive components such as the relationships between elements unfolding through time, arranging the embedding space to emphasize the clustering of these sophisticated musical content without considering sonic qualities or attributes.

### 2023-06-08 introduction/Subsections/Previous work.tex
Updates from Overleaf

Recent advancements in neural networks have enabled the development of low-dimensional embeddings which encapsulate critical musical attributes, thus facilitating the computation of diverse task-specific elements.

Such tasks encompass developing an audio embedding method that excels across a wide array of applications without fine-tuning \cite{Turian2022HEAR:Representations}, deploying deep audio embeddings for multi-level music segmentation \cite{SalamonDeepSegmentation}, improving environmental sound classification \cite{Kim2020OneStrategies}, enhancing vocal-centric music tasks via cross-domain audio embedding \cite{Kim2021LearningLoss}, combining task-specific and pre-trained features for improved audio classification \cite{Hung2022Feature-informedClassification}, creating a music similarity search engine tailored for video producers \cite{epidemic}, enhancing Music Emotion Recognition (MER) performance \cite{KohComparisonRecognition}, investigating the efficacy of speaker recognition pre-trained model embeddings \cite{lightweight}, embedding songs for similarity comparison via artist identification \cite{contentmusicsimtriplet2020}, solving the cross-modal text-to-music retrieval issue \cite{WonEmotionStories}, and facilitating automated music rearrangement \cite{Stoller2018IntuitiveTransitions, Plachouras2023MusicSegmentation}.

Furthermore, deep learning applications in music often overlook the intermediate-level musical patterns, analogous to the 'mesostructure' in music and the middle ground in Schenkerian analysis. As highlighted by \cite{Mesostructures2023}, addressing this mesostructure could significantly improve tasks like music analysis, composition, and retrieval by providing a more holistic understanding of music, comparable to insights from Schenkerian analysis \cite{Introduction_to_Schenkerian_Analysis}.

@@

However, it is still an open question whether a general-purpose audio representation can effectively mimic human hearing \cite{Turian2022HEAR:Representations}.

Considering their proven efficacy in existing literature and their potential for transfer learning, it appears promising to further investigate deep audio embeddings for Music Information Retrieval (MIR) downstream tasks.

>>>>

Prior research has underlined the importance of choosing audio features for effectively identifying and labeling distinct music segments in music segmentation. However, acquiring music segment annotation data for learning feature transformations has been a limiting factor due to its time-consuming and costly nature. To overcome this, studies have employed unsupervised deep learning, using abundantly available music audio. Combining spectrograms with self-similarity lag matrices has captured a broader spectrum of structural information. This approach has significantly improved the performance of segmentation algorithms, achieving state-of-the-art results. Additionally, strategies considering the hierarchical nature of the structural organization, utilizing two-level annotations of main and secondary boundaries, have further enhanced boundary recognition performance \cite{unsupervisedlearndeepfeat, GrillMUSICANNOTATIONS, Hernandez-Olivan2021MusicFeatures, SalamonDeepSegmentation}.

Aside from the music structure applications, recent advancements in neural networks have enabled the development of low-dimensional embeddings which encapsulate critical musical attributes, thus facilitating the computation of diverse task-specific elements. Such tasks encompass developing an audio embedding method that excels across a wide array of applications without fine-tuning \cite{Turian2022HEAR:Representations}, improving environmental \cite{Kim2020OneStrategies} sound classification \cite{CramerLOOKEMBEDDINGS}, enhancing vocal-centric music tasks via cross-domain audio embedding \cite{Kim2021LearningLoss}, combining task-specific and pre-trained features for improved audio classification \cite{Hung2022Feature-informedClassification}, creating a music similarity search engine tailored for video producers \cite{epidemic}, enhancing Music Emotion Recognition (MER) performance \cite{KohComparisonRecognition}, investigating the efficacy of speaker recognition pre-trained model embeddings \cite{lightweight}, embedding songs for similarity comparison via artist identification \cite{contentmusicsimtriplet2020}, solving the cross-modal text-to-music retrieval issue \cite{WonEmotionStories}, and facilitating automated music rearrangement \cite{Stoller2018IntuitiveTransitions, Plachouras2023MusicSegmentation}.

@@

Considering their proven promising efficacy in the existing literature and their potential for transfer learning, it appears appealing to investigate further deep audio embeddings for MIR downstream tasks. Especially as it is still an open question whether a general-purpose audio representation can effectively mimic human hearing \cite{Turian2022HEAR:Representations} even though some approaches achieved to generalize up to fourteen music understanding tasks \cite{Li2023MERT:Training}.

### 2023-06-08 methods/subsections/implementation.tex
Updates from Overleaf

Contrastive Learning of Musical Representations (CLMR) learns valuable, discriminative music representations without explicit labels by contrasting positive augmentations of a musical piece against negative ones \cite{CLMR2021}. CLMR works under a machine learning subset called Self-Supervised Learning (SSL), where models self-learn from unlabeled data, creating their supervisory signal \cite{audioselfsupsurvey}. This mimics human learning from observations and interactions, transforming unsupervised problems into supervised ones by auto-generating labels. The benefits of SSL include reduced reliance on labeled data, making data representation more robust and generalizable.

@@

@@

\subsubsection{Learning rate}

@@

\subsubsection{Batch normalization}

>>>>

Contrastive Learning of Musical Representations (CLMR) learns valuable, discriminative music representations without explicit labels by contrasting positive augmentations of a musical piece against negative ones \cite{CLMR2021}. CLMR works under a machine learning subset called self-supervised learning (SSL), where models self-learn from unlabeled data, creating their supervisory signal \cite{audioselfsupsurvey}. This mimics human learning from observations and interactions, transforming unsupervised problems into supervised ones by auto-generating labels. The benefits of SSL include reduced reliance on labeled data, making data representation more robust and generalizable.

@@

Although SSL has been proven effective in speech and audio, its
application to music audio has yet to be thoroughly explored. This is primarily due
to the distinctive challenges associated with modeling musical knowledge, particularly tonal and pitched characteristics of music \cite{Li2023MERT:Training}.

@@

@@

### 2023-06-08 results/subsections/tables.tex
Updates from Overleaf

    \item \textbf{F-measure}: The F-measure is a widely used boundary detection metric computed by comparing the predicted and ground truth boundaries. This metric, ranging from 0 to 1, represents the harmonic mean of precision and recall, providing a balanced view of under-segmentation and over-segmentation. Precision calculates the percentage of correctly detected boundaries, whereas recall assesses the percentage of actual boundaries detected correctly \cite{Turnbull2007ABOOSTING}. The F-measure is calculated as follows:

@@

\begin{table}[h]
\centering
\small
\begin{tabularx}{\textwidth}{>{\raggedright\arraybackslash}p{4.5cm}XXXXX}
\toprule
\thead{\centering\textbf{Year, Authors, Ref}} & \thead{\centering\textbf{Algorithm}} & \thead{\centering\textbf{Input}} & \thead{\centering\textbf{Method}} & \thead{\centering\textbf{F-measure}} \\
\midrule
\addlinespace
2012, Kaiser et al., [27]& KSP2 & SSM & Novelty measure  & 0.286 \\
\addlinespace
2013, McFee \& Ellis, [20] & MP2 & MLS & Fisher’s Linear Discriminant  & 0.317 \\
\addlinespace
2014, Nieto \& Bello, [28] & NB1 & MFCCs, chromas & Checkerboard-like kernel  & 0.299 \\
\addlinespace
2015, Cannam et al., [29] & CC1 & Timbre-type histograms & HMM  & 0.213 \\
\addlinespace
2016, Nieto, [30] & ON2 & Constant-Q Transform Spectrogram & Linear Discriminant Analysis  & 0.299 \\
\addlinespace
2017, Cannam et al., [29] & CC1 & Timbre-type histograms & HMM  & 0.212 \\
\addlinespace
2007, Turnbull et al., [19] & - & MFCCs, chromas, spectrogram & Boosted Decision Stump  & 0.378 \\
\addlinespace
2011, Sargent et al., [34] & - & MFCCs, chromas & Viterbi  & 0.356 \\
\addlinespace
2014, Ullrich et. al, [22] & - & MLS & CNN  & 0.465 \\
\addlinespace
2015, Grill \& Schlüter, [4] & - & MLS, SSLMs & CNN  & 0.523 \\
\addlinespace
2015, Grill \& Schlüter, [5] & - & MLS, PCPs, SSLMs & CNN  & 0.508 \\
\addlinespace
2017, Hadria \& Peeters, [35] & - & MLS, SSLMs & CNN  & 0.291 \\
\bottomrule
\end{tabularx}
\caption[Baseline. State of the art table.]{Your Caption. Given the potential inaccuracies in human annotations and prediction errors, the F-measure tolerates minor discrepancies between predicted and ground truth boundaries. This tolerance level can be adjusted, enabling the metric to consider a predicted boundary as correct if it's within the set tolerance window of a ground truth boundary.}
\label{tab:my_label}
\end{table}

>>>>

    \item \textbf{F-measure}: The F-measure is a widely used boundary detection metric computed by comparing the predicted and ground truth boundaries. This metric, ranging from 0 to 1, represents the harmonic mean of precision and recall, providing a balanced view of under-segmentation and over-segmentation. Given the potential inaccuracies in human annotations and prediction errors, the F-measure tolerates minor discrepancies between predicted and ground truth boundaries. This tolerance level can be adjusted, enabling the metric to consider a predicted boundary as correct if it's within the set tolerance window of a ground truth boundary \cite{Turnbull2007ABOOSTING}. The F-measure is calculated as follows:

@@

\begin{table}[h]
\centering
\small
\begin{tabularx}{\textwidth}{>{\raggedright\arraybackslash}p{4.5cm}XXXXX}
\toprule
\thead{\centering\textbf{Year, Authors, Ref}} & \thead{\centering\textbf{Algorithm}} & \thead{\centering\textbf{Input}} & \thead{\centering\textbf{Method}} & \thead{\centering\textbf{F-measure}} \\
\midrule
\addlinespace
2012, Kaiser et al., [27]& KSP2 & SSM & Novelty measure  & 0.286 \\
\addlinespace
2013, McFee \& Ellis, [20] & MP2 & MLS & Fisher’s Linear Discriminant  & 0.317 \\
\addlinespace
2014, Nieto \& Bello, [28] & NB1 & MFCCs, chromas & Checkerboard-like kernel  & 0.299 \\
\addlinespace
2015, Cannam et al., [29] & CC1 & Timbre-type histograms & HMM  & 0.213 \\
\addlinespace
2016, Nieto, [30] & ON2 & Constant-Q Transform Spectrogram & Linear Discriminant Analysis  & 0.299 \\
\addlinespace
2017, Cannam et al., [29] & CC1 & Timbre-type histograms & HMM  & 0.212 \\
\addlinespace
2007, Turnbull et al., [19] & - & MFCCs, chromas, spectrogram & Boosted Decision Stump  & 0.378 \\
\addlinespace
2011, Sargent et al., [34] & - & MFCCs, chromas & Viterbi  & 0.356 \\
\addlinespace
2014, Ullrich et. al, [22] & - & MLS & CNN  & 0.465 \\
\addlinespace
2015, Grill \& Schlüter, [4] & - & MLS, SSLMs & CNN  & 0.523 \\
\addlinespace
2015, Grill \& Schlüter, [5] & - & MLS, PCPs, SSLMs & CNN  & 0.508 \\
\addlinespace
2017, Hadria \& Peeters, [35] & - & MLS, SSLMs & CNN  & 0.291 \\
\bottomrule
\end{tabularx}
\caption[Baseline. State of the art table.]{Your Caption.}
\label{tab:my_label}
\end{table}
